---
title: "Dropseq analysis"
date: "`r Sys.Date()`"
output: 
  flexdashboard::flex_dashboard
---


```{r, echo=FALSE, cache=FALSE}
## Define the sample prefix used for all qc files as well as the targeted number of STAMPS
sample_prefix <- ""
number_of_STAMPS=2000
```


```{r knitr_init, echo=FALSE}

## Will check for the required R packages and quietly install them if necessary

library(knitr)
library(rmdformats)
library(ggplot2)
library(reshape)
library(pander)
library(gtools)
library(gridExtra)
library(dplyr)
library(ggthemes)
library(plotly)

##Plotting theme from (https://rpubs.com/Koundy/71792)
theme_Publication <- function(base_size=14, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family=base_family)
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

```

**Barcode quality**{data-orientation=rows}
===================================== 

### **Cell barcodes**

```{r,echo=FALSE}
cell_barcode_name <- paste(sample_prefix,"unaligned_tagged_Cellular.bam_summary.txt",sep=".")
cell_barcode <- read.table(paste("./qc_stats/",cell_barcode_name,sep=""),sep="\t",header=T)

cell_barcode$num_failed_bases <- as.character(cell_barcode$num_failed_bases)
cell_barcode$num_failed_bases <- factor(cell_barcode$num_failed_bases,levels=mixedsort(cell_barcode$num_failed_bases))

number_cell_barcodes <- sum(cell_barcode$num_barcodes)

plot_cell_barcode <- ggplot(cell_barcode,aes(x = num_failed_bases, y= log10(num_barcodes))) +
  geom_bar(fill="#db4437",stat="identity") +
  labs(x="Number of failed bases" , y="Number of log10(Barcodes)") +
  theme_bw() +
  theme(legend.position="none") +
  theme_Publication()

plot_cell_barcode
```

### **Molecular barcodes**

```{r,echo=FALSE}
molecular_barcode_name <- paste(sample_prefix,"unaligned_tagged_Molecular.bam_summary.txt",sep=".")
molecular_barcode <- read.table(paste("./qc_stats/",molecular_barcode_name,sep=""),sep="\t",header=T)
molecular_barcode$num_failed_bases <- as.character(molecular_barcode$num_failed_bases)
plot_molecular_barcode <- ggplot(molecular_barcode,aes(x = num_failed_bases, y= log10(num_barcodes))) +
  geom_bar(fill="#4285f4",stat="identity") +
  labs(x="Number of failed bases" , y="Number of log10(Barcodes)") +
  theme_bw() +
  theme(legend.position="none") +
  theme_Publication()

plot_molecular_barcode
```

**Adapter & polyA trimming**{data-orientation=rows}
===================================== 

### **Adapter trimming distribution**

```{r,echo=FALSE}
adapter_report_name <- paste(sample_prefix,"adapter_trimming_report.txt",sep=".")
adapter_report <- read.table(paste("./qc_stats/",adapter_report_name,sep=""),
                             skip=4,
                             sep="\t",
                             header=T)

all_reads_trimmed <- sum(adapter_report$VALUE)

plot_adapter_report <- ggplot(adapter_report,aes(x = BIN, y= VALUE)) +
  geom_bar(fill="#0f9d58",stat="identity") +
  labs(x="BIN" , y="Value") +
  theme_bw() +
  theme(legend.position="none") +
  theme_Publication()
plot_adapter_report
```


### **PolyA trimming distribution**

```{r,echo=FALSE}
polyA_report_name <- paste(sample_prefix,"polyA_trimming_report.txt",sep=".")
polyA_report <- read.table(paste("./qc_stats/",polyA_report_name,sep=""),
                             skip=4,
                             sep="\t",
                             header=T)

polyA_trimmed <- sum(polyA_report$VALUE)

plot_polyA_report <- ggplot(polyA_report,aes(x = BIN, y= VALUE)) +
  geom_bar(stat="identity") +
  labs(x="BIN" , y="Value") +
  theme_bw() +
  theme(legend.position="none") +
  theme_Publication()
plot_polyA_report
```

**Barcode Synthesis errors** {data-orientation=rows}
===================================== 

### **Barcode Synthesis error types**

```{r,echo=FALSE}
synthesis_name <- paste(sample_prefix,"synthesis.summary.txt",sep=".")

synthesis_types<- read.table(paste("./qc_stats/",synthesis_name,sep=""),
                             skip=1,
                             nrows=1,
                             sep="\t",
                             header=T)

synthesis_types_long <- melt(synthesis_types)

plot_synthesis_types<- ggplot(synthesis_types_long, aes(x=variable,y= value,fill=variable)) +
  geom_bar(stat="identity") +
  labs(x="Synthesis Error Type" , y="Number of Barcodes") +
  theme_bw() +
  theme(legend.position="none") +
  coord_flip() +
  scale_fill_brewer(palette="Set1")

plot_synthesis_types

```

### **Synthesis error base distribution**

```{r,echo=FALSE}
synthesis_distribution <- read.table(paste("./qc_stats/",synthesis_name,sep=""),
                             skip=4,
                             sep="\t",
                             header=T)

synthesis_distribution$SYNTHESIS_ERROR_BASE <- as.character(synthesis_distribution$SYNTHESIS_ERROR_BASE)

plot_synthesisdistribution <- ggplot(synthesis_distribution,aes(x = SYNTHESIS_ERROR_BASE, y= num.cells)) +
  geom_bar(stat="identity") +
  labs(x="Synthesis Error Base" , y="Number of Cells") +
  theme_Publication()

plot_synthesisdistribution

```



**UMI statistics**{data-orientation=rows}
===================================== 

Column {data-width=600}
-------------------------------------
### **Duplicated UMIs vs unique UMIs per Cell**

```{r,echo=FALSE}
library(stringr)
## Look at the number of duplicated and unique UMIs per Cell

umi_by_gene <- read.table("./qc_stats/UMI_by_gene_dist.tab",sep="\t",header=T)

## Count number of duplicated UMIs per cell
num_of_duplicated_umis_per_cell <- umi_by_gene %>%
  group_by(Cell.Barcode) %>%
  subset(Num_Obs > 1) %>%
  count()
colnames(num_of_duplicated_umis_per_cell) <- c("Cell.Barcode","Num_duplicated_UMIs")

## Count number of unique UMIs per cell
unique_umis_per_cell <- umi_by_gene %>%
  group_by(Cell.Barcode) %>%
  subset(Num_Obs == 1) %>%
  count()
colnames(unique_umis_per_cell) <- c("Cell.Barcode","Num_of_unique_UMIs")

## Sum up all duplicated UMIs
subset_duplicated_umis <- umi_by_gene %>%
  group_by(Cell.Barcode) %>%
  subset(Num_Obs > 1) %>%
  summarise(Mean_of_duplications_per_UMI=mean(Num_Obs))
  
subset_duplicated_umis_sum <- umi_by_gene %>%
  group_by(Cell.Barcode) %>%
  subset(Num_Obs > 1) %>%
  summarise(Sum_of_duplications_per_UMI=sum(Num_Obs))

## Join duplicated and unique counts into one tibble
joined_umi_counts <- full_join(num_of_duplicated_umis_per_cell,unique_umis_per_cell,by="Cell.Barcode")
joined_umi_counts <- full_join(joined_umi_counts,subset_duplicated_umis,by="Cell.Barcode")
joined_umi_counts <- full_join(joined_umi_counts,subset_duplicated_umis_sum,by="Cell.Barcode")

## Calculate the total number of UMIs as well as the percent of duplicated UMIs
joined_umi_counts <- joined_umi_counts %>%
  group_by(Cell.Barcode) %>%
  mutate(Total_number_individual_UMIs_detected = sum(Num_duplicated_UMIs,Num_of_unique_UMIs)) %>%
  mutate(Percent_duplicate_UMIs = Num_duplicated_UMIs/Total_number_individual_UMIs_detected*100) %>%
  mutate(Percent_unique_UMIs = Num_of_unique_UMIs/Total_number_individual_UMIs_detected*100)

## Sort table based on Total number of UMIs per Cells
joined_umi_counts <- joined_umi_counts %>%
  arrange(desc(Percent_unique_UMIs))

## Calculate GC content for Cell barcodes
joined_umi_counts <-  joined_umi_counts %>%
  mutate(count_G= str_count(Cell.Barcode, "G")) %>%
  mutate(count_C= str_count(Cell.Barcode, "C")) %>%
  mutate(freq_GC= (count_G + count_C) / 12)

joined_umi_counts <- joined_umi_counts %>%
  arrange(desc(freq_GC))

## Plot correlation between duplicated and unique UMIs per cell
mean_duplicated_UMI <- ggplot(joined_umi_counts,aes(Num_duplicated_UMIs,Num_of_unique_UMIs,colour=Total_number_individual_UMIs_detected)) +
  geom_point(aes(size=Mean_of_duplications_per_UMI),fill="grey50") +
  theme_light() +
  labs(title="Duplicated vs unique UMIs per Cell",
       x = "Number of duplicated UMIs",
       y = "Number of unique UMIs",
       subtitle = "Every point represents a Cell! Coloured by Percentage of duplicated UMIs for Cell barcode") +
  theme(plot.title = element_text(lineheight=.8, face="bold"))

ggplotly(mean_duplicated_UMI)

```

### **Duplicated UMIs vs unique UMIs per Gene**

```{r}
## Look at the number of duplicated and unique UMIs per Gene
## Count number of duplicated UMIs per cell
num_of_duplicated_umis_per_gene <- umi_by_gene %>%
  group_by(Gene) %>%
  subset(Num_Obs > 1) %>%
  count()
colnames(num_of_duplicated_umis_per_gene) <- c("Gene","Num_duplicated_UMIs")

## Count number of unique UMIs per cell
unique_umis_per_gene <- umi_by_gene %>%
  group_by(Gene) %>%
  subset(Num_Obs == 1) %>%
  count()
colnames(unique_umis_per_gene) <- c("Gene","Num_of_unique_UMIs")

## Join duplicated and unique counts into one tibble
joined_umi_gene_counts <- full_join(num_of_duplicated_umis_per_gene,unique_umis_per_gene,by="Gene")

## Calculate the total number of UMIs as well as the percent of duplicated UMIs
joined_umi_gene_counts <- joined_umi_gene_counts %>%
  group_by(Gene) %>%
  mutate(Total_number_individual_UMIs_detected = sum(Num_duplicated_UMIs,Num_of_unique_UMIs)) %>%
  mutate(Percent_duplicate_UMIs = Num_duplicated_UMIs/Total_number_individual_UMIs_detected*100) %>%
  mutate(Percent_unique_UMIs = Num_of_unique_UMIs/Total_number_individual_UMIs_detected*100)

## Sort table based on Total number of UMIs per Cells
joined_umi_gene_counts <- joined_umi_gene_counts %>%
  arrange(desc(Num_of_unique_UMIs))

umi_duplication_by_gene <- ggplot(joined_umi_gene_counts,aes(Num_duplicated_UMIs,Num_of_unique_UMIs,colour=Percent_duplicate_UMIs,text=Gene)) +
  geom_point(size=2,fill="grey50") +
  theme_light() +
  labs(title="Duplicated vs Unique UMIs per Gene",
       x = "Number of duplicated UMIs",
       y = "Number of unique UMIs",
       subtitle = "Every represents is a Gene!") +
  theme(plot.title = element_text(lineheight=.8, face="bold"))

ggplotly(umi_duplication_by_gene)
```

Column {data-width=600}
-------------------------------------

### **Distribution of duplicated UMIs per cell**

```{r,echo=FALSE}
ggplot(joined_umi_counts,aes(log10(Mean_of_duplications_per_UMI))) +
  geom_histogram(colour="black",fill="grey50") +
  theme_light() +
  labs(title="Distribution of mean UMI duplication per Cell",
       subtitle = "For all duplicated UMIs, mean of duplication events was calculated per Cell.",
       x = "log10 of mean UMI duplications events per Cell)",
       y = "Number of cells") +
  theme(plot.title = element_text(lineheight=.8, face="bold"))
```


### **Distribution of duplicate UMIs per cell**

```{r,echo=FALSE}
ggplot(joined_umi_counts,aes(Percent_duplicate_UMIs)) +
  geom_histogram(colour="black",fill="grey50") +
  theme_light() +
  labs(title="Percentage of duplicated UMIs per cell",
       subtitle = "Distribution of percentage of duplicated UMIs",
       x = "Duplicate UMIs (%)",
       y = "Number of cells") +
  theme(plot.title = element_text(lineheight=.8, face="bold"))
```


**Read Distributions**{data-icon="ion-connection-bars"}
===================================== 

### **Cumulative Read distribution**

```{r,echo=FALSE}
read_count_distribution_name <- paste(sample_prefix,"readcount.txt",sep=".")
read_count_distribution <- read.table(paste("./qc_stats/",read_count_distribution_name,sep=""),
                             sep="\t",
                             header=F,
                             stringsAsFactors=F)

cumulative_sum =cumsum(read_count_distribution$V1)
cumulative_sum=cumulative_sum/max(cumulative_sum)

STAMPS_barcodes <- read_count_distribution[1:number_of_STAMPS,2]
write.table(file="STAMP_barcodes.txt",
            STAMPS_barcodes,
            row.names=FALSE,
            col.names=FALSE,
            sep="\t",
            quote=FALSE)

x_cutoff=number_of_STAMPS*50
cumulative_sum_matrix <- data.frame("Value"=cumulative_sum,"Cell_number"=1:length(cumulative_sum))
cumulative_sum_matrix_subset <- subset(cumulative_sum_matrix,Value < x_cutoff)
read_count_distribution$Value <- 1:nrow(read_count_distribution)

plot_cumulative_sum <- ggplot(cumulative_sum_matrix_subset,aes(x=Cell_number, y=Value)) +
  geom_line() +
  scale_x_continuous(limits=c(0,x_cutoff)) +
  geom_vline(colour="red",xintercept=number_of_STAMPS) +
  labs(x="cell barcodes sorted by number of reads [descending]",
       y="cumulative fraction of reads") +
  theme_Publication()
ggplotly(plot_cumulative_sum)
```


### **Read distribution for top barcodes**

```{r,echo=FALSE}
## Total number of reads
all_reads <- sum(read_count_distribution$V1)

## Plot the number of barcodes for the defined number of barcdoes sequenced
read_count_distribution_subset <- read_count_distribution[1:number_of_STAMPS,]
read_count_distribution_subset$V1 <- as.numeric(read_count_distribution_subset$V1)
plot_read_count_distribution_subset <- ggplot(read_count_distribution_subset,aes(log10(V1))) +
  geom_histogram() +
  labs(x="Log10(Number of reads)",
       y="Number of STAMPS",
       title="Distribution of Reads for the top 2000 STAMPS") +
  theme_Publication()
ggplotly(plot_read_count_distribution_subset)
```
