---
title: "Dropseq_coverage_estimator"
author: "Florian Wuennemann"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dropseq depth coverage estimator for STAMPS

This utility lets you explore how many reads you expect per STAMP based on multiplexing your libraries on different illumina machines and using different amounts of STAMPS. There are several parameters that have to be defined which are rough estimates and are dependent on your specific Dropseq run. The default values are set to reasonable estimations but might not apply to your specific run.

```{r,  echo = FALSE, eval=FALSE}
reads <- 4e8
cumulative <- 0.5
number_of_STAMPS <- 2000

distribution <- rnbinom(2000,0.7,0.5)
distribution_sorted <- sort(distribution,decreasing=TRUE)

cumulative_sum =cumsum(distribution_sorted)
cumulative_sum=cumulative_sum/max(cumulative_sum)

cumulative_sum_matrix <- data.frame("Value"=cumulative_sum,"Cell_number"=1:length(cumulative_sum))

```




```{r, echo = FALSE}
library(ggplot2)
shinyApp(
  
  ui = fluidPage(
    ## Cumulative read distribution
    sliderInput(inputId="cum_read_dist", label=h3("Cumulative Read Distribution"), min=0,max=1,value=0.5),
     ## Number of reads going to library
    sliderInput(inputId="num_of_stamps", label=h3("Number of STAMPS"), min=0,max=10000,value=2000),
    
    plotOutput("Cumulative_plot"),
    
    plotOutput("Distribution_plot")
  ),
  
  server = function(input, output) {
  
    distribution <- reactive({
      dist <- rnbinom(input$num_of_stamps,2,input$cum_read_dist)
      dist <- sort(dist,decreasing=TRUE)
      data.frame("Value"=dist,"Cell_number"=1:length(dist))
    })
    
    cumulative_sum_matrix <- reactive({ 
      
      cumulative_sum <- cumsum(distribution()$Value)
      cumulative_sum <- cumulative_sum/max(cumulative_sum)

      data.frame("Value"=cumulative_sum,"Cell_number"=1:length(cumulative_sum))
    })
    
    mean_depth <- reactive ({
      reads <- subset(distribution()$V1[2000,])
      mean(reads)
    })

    output$Cumulative_plot <- renderPlot({
      plot_cumulative_sum <- ggplot(cumulative_sum_matrix(),aes(x=Cell_number, y=Value)) +
        geom_line() +
        scale_x_continuous() +
        labs(x="cell barcodes sorted by number of reads [descending]",
             y="cumulative fraction of reads") +
        theme_light()
      print(plot_cumulative_sum)
    })
    
    output$Distribution_plot <- renderPlot({
      plot_distribution <- ggplot(distribution(),aes(x=Cell_number, y=Value)) +
        geom_line() +
        scale_x_continuous() +
        labs(x="cell barcodes sorted by number of reads [descending]",
             y="cumulative fraction of reads") +
        theme_light()
      print(plot_distribution)
    })
    
  },
  
  options = list(height = 2000)
)
```